### 2014-10-08 paper/include/background.tex
wrote stuff

To refresh one rank, the simplest approach is to refresh all banks in parallel row-by-row at the start of the 64 ms interval. This approach is called Burst refresh and has some downsides. The approach blocks normal accesses for a longer period and form a peak of power consumption during the burst time. A better alternative is to distribute the refreshes across the interval, resulting in a lower average latency for the processor and a more stable power consumption.

@@

The conclusion drawn from \reffig{fig:dram_data_proj} is that it will be increasingly important to mitigate the throughput loss and DRAM power consumption of future systems. One way to decrease these negative trends of DRAM scaling is to optimize the DRAM refresh.

>>>>

To refresh one rank, the simplest approach is to refresh all banks in parallel row-by-row at the start of the 64 ms interval. This approach is called burst refresh and has some downsides. The approach blocks normal accesses for a longer period and form a peak of power consumption during the burst time. A better alternative is to distribute the refreshes across the interval, resulting in a lower average latency for the processor and a more stable power consumption.

@@

The conclusion drawn from \reffig{fig:dram_data_proj} is that it will be increasingly important to mitigate the throughput loss and DRAM power consumption of future systems. One way to decrease these negative trends of DRAM scaling is to optimize the DRAM refresh. However, a reduction of the refreshes made has no linear relationship to the reduction of refresh energy consumption. The energy reduction depends on the state of the bank in which the targeted row resides; if the targeted row is closed more energy will be consumed than if the targeted row would have been open.

### 2014-10-08 paper/include/refrint.tex
more reorganization
made some placeholders for new subsections
wrote stuff

@@

\subsection{Differences and similarities}

>>>>

This technique is proposed by A. Agrawal et al. \cite{refrint} and is based on Access Recency. It is targeted towards caches and eDRAM and thus can not be directly compared with the other solutions, but the main concept is of good use and the technique could be implemented DRAM as well. The concept is based on two types of unnecessary refreshes which originates from so called cold and hot rows. Cold rows are those used far apart in time or not used at all. The technique identifies and refreshes only those rows which are expected to be accessed in the near future, the rows that are used less frequent is flushed and invalidated. Hot rows consists of the rows that gets accessed frequently and thus can the technique postpone the preceding refresh to the accessed rows.

For the hot lines, an approach similar to Smart Refresh is employed. The differences are that Refrint does not maintain a counter for each row, but a snapshot of a global counter.

\todo[inline]{Crunchable text below!}

@@

\subsubsection*{Differences and similarities}

### 2014-10-08 paper/include/smart_refresh.tex
wrote stuff

The basic concept behind our scheme is that a memory line that has been recently read out or written to does not need to be refreshed again by the periodic refresh mechanism.

Uses a time-out counter for each memory line. The counter is set on a read/write/refresh and dicards refreshes until the timer hits zero. The counter is associated to a bank/line pair. The counters are stored in the memory controller. One counter consists of 2 SRAM bits. More bits give higher granularity and thus also better performance because the refresh is made less often.

This scheme will be most effective if all lines are accessed before a refresh is needed. If no accesses are made = the entire working set fits into cache, Smart Refresh becomes CBR (CAS before RAS). Only refresh one line at a time = inefficient, so they have made it possible to turn the technique off.

Implemented so only one line per each logical segments are deremented at the same time, which leads to that the refreshes are evenly distributed over time. (I.e. no burst refresh.)

Overhead: 48KB for 2GB and 768 KB for 32GB = 0,0024\% overhead.

Average reduction of refreshes by 59.3\% for 2GB and 52.3\% energy saving regarding the DRAM refresh. No performance degradation.

Employed its solution to 2 GB ram. What about todays sizes? How effective is the technique now? - Only works effective if large parts the memory is acccessed, i.e. large part of the memory has to be valid to begin with. Is it so today?

>>>>

M. Ghosh and H.-H. S. Lee proposes the technique Smart Refresh \cite{smartrefresh} that targets Access Recency. The basic idea is to maintain a time-out counter for each memory row and when a counter time-out, the MC issues a ROR refresh to the corresponding row. The counters are set if the row is accessed or refreshed.

The counters are decremented in such a fashion so the refreshes are distributed evenly over time and thus avoiding burst refresh. To achive this, the counters are initially staggered and divided into n logical blocks where the counters are accessed in a sequential order. Only n counters will be accessed, set or decremented at the same point in time, which result in that maximum n rows are refreshed simultainously.

The counters bit size corresponds to the time granularity in where an access can be captured. A higher bit size results in that an access can be captured later in time, which effectively postpones the refresh further. The least optimal case for postponing the refresh is if the row access occurs a small time intervall before the counter is accessed. If a 2-bit counter is used the least optimal row access result in that the row is refreshed after 3/4 of the maximum refresh period. Analogously, a 3-bit counter result in that a row is refreshed after minimum 7/8 of the maximum refresh period.

If the row accesses drops below 1\% during a maximum refresh period, Smart Refresh is disabled. As stated, an access recency technique is dependent on row accesses to be of gain. When disabled, the MC uses it's default CBR refresh scheme. Smart refresh is then activated if the row accesses increases above 2\%.

Smart Refresh modifies the MC but does not change the DRAM module nor the interface between them. The modifications of the MC constitutes of counter update logic, the counter array, and a pending refresh request queue. The memory footprint of the modifications is 0.0024\% of total device capacity when 3-bits counters are used.

The technique result in an average refresh reduction of 59\% and saves on average 52\% of refresh power consumption without any performance degradation. The simulation was performed on 2 GB of main memory.
